CCD integration experiment - Motorcycle study with Gaussian processes

```{r}
source('CCD_point_generation.r')
library(expm)
library(ggplot2)
library(latex2exp)
library(tidyverse)
library(cmdstanr)
library(posterior)
theme_set(bayesplot::theme_default(base_family = "sans", base_size=16))
set1 <- RColorBrewer::brewer.pal(7, "Set1")
```

# Full data set experiment

## Obtaining the data

```{r}
library(rjson)
library(MASS)
data(mcycle, package="MASS")

# full data set
standata_gpcovf <- list(x=mcycle$times,
                        x2=mcycle$times,
                        y=mcycle$accel,
                        N=length(mcycle$times),
                        N2=length(mcycle$times))
json_file <- toJSON(standata_gpcovf)
write(json_file, "gpcovf.json")
```

## MCMC sampling with cmdstanr

```{r}
model_gpcovf <- cmdstan_model(stan_file = "gpcovf.stan")
fit_gpcovf <- model_gpcovf$sample(data=standata_gpcovf,
                                  iter_warmup=500, iter_sampling=1000,
                                  chains=4, parallel_chains=4, refresh=1000)
```

```{r}
draws_gpcovf <- as_draws_rvars(fit_gpcovf$draws())
summarise_draws(subset(draws_gpcovf,
                       variable=c('sigma_','lengthscale_','sigma'),
                       regex=TRUE))
```


## Fitting the model with bridgestan

This model is used only to enable the posterior evaluations needed when running the CCD point generation algorithm.

```{r}
library(bridgestan)
model <- StanModel$new("gpcovf_model.so", "gpcovf.json", 1234)
```


```{r}
print(paste0("This model's name is ", model$name(), "."))
print(paste0("This model has ", model$param_num(), " parameters."))
print(paste0("This model has parameters [", toString(model$param_unc_names()), "]."))
```

## Conducting the CCD point generation

```{r}
logdensity <- function(theta) {
  model$log_density(theta, jacobian=TRUE)
}
logdensity_gradient <- function(theta) {
  model$log_density_gradient(theta,  jacobian = TRUE)$gradient
}
logdensity_hessian <- function(theta) {
  model$log_density_hessian(theta,  jacobian = TRUE)$hessian
}

source('CCD_point_generation.r')
ccd <- CCD(logdensity,
           3,
           th0=c(-1,0,-1),
           logden.grad=logdensity_gradient,
           logden.hessian=logdensity_hessian,
           optim.method='BFGS',
           auto.scale = T)
```

## Some visualization for validation
```{r, fig.width=5}
xl <- c(0.1,0.65)
yl <- c(0,3)
icons <- c(rep(22, 8), rep(21, 7))
colors <- c(rep('steelblue', 14), 'red')

as.data.frame(exp(ccd$points)) %>% map_df(rev) %>%
  ggplot(aes(x=V1, y=V2)) +
    coord_cartesian(xlim = xl, ylim = yl) +
    labs(x=TeX(r"(lengthscale $l_p$)"), y=TeX(r"($\sigma^2_f$)")) +
    geom_point(data=as_draws_df(draws_gpcovf), aes(x=lengthscale_f, y=sigma_f), col=alpha('darkgrey', 0.15), pch=19) +
    geom_point(pch=icons, col=colors, size=2, stroke=1) +
    theme_bw() 
    # + theme(axis.title.y = element_text(angle = 0))

xl <- c(0.35,0.65)
as.data.frame(exp(ccd$points)) %>% map_df(rev) %>%
  ggplot(aes(x=V3, y=V2)) +
    coord_cartesian(xlim = xl, ylim = yl) +
    labs(x=TeX(r"($\sigma_n$)"), y=TeX(r"($\sigma^2_f$)")) +
    geom_point(data=as_draws_df(draws_gpcovf), aes(x=sigman, y=sigma_f), col=alpha('darkgrey', 0.15), pch=19) +
    geom_point(pch=icons, col=colors, size=2, stroke=1) +
    theme_bw()

```

## Optimized model

```{r}
library(posterior)
opt_gpcovf <- model_gpcovf$optimize(data=standata_gpcovf,
                                    init=0.1, algorithm='bfgs')
odraws_gpcovf <- as_draws_rvars(opt_gpcovf$draws())
mode <- c(E(odraws_gpcovf$lengthscale_f), E(odraws_gpcovf$sigma_f), E(odraws_gpcovf$sigman))
subset(odraws_gpcovf, variable=c('sigma_','lengthscale_','sigma'), regex=TRUE)

```

## Predictive plots

### Optimized model

```{r}
mcycle %>%
  ggplot(aes(x=times,y=accel))+
  geom_point()+
  labs(x="Time (ms)", y="Acceleration (g)", title="")+
  geom_line(data=mcycle,aes(x=times,y=mean(odraws_gpcovf$f)), color=set1[1])+
  geom_line(data=mcycle,aes(x=times,y=mean(odraws_gpcovf$f-2*odraws_gpcovf$sigma)), color=set1[1],
            linetype="dashed")+
  geom_line(data=mcycle,aes(x=times,y=mean(odraws_gpcovf$f+2*odraws_gpcovf$sigma)), color=set1[1],
            linetype="dashed")
```

### CCD-based model

```{r}
mcycle %>%
  ggplot(aes(x=times,y=accel))+
  geom_point()+
  labs(x="Time (ms)", y="Acceleration (g)", title="")+
  geom_line(data=mcycle,aes(x=times,y=mean(odraws_gpcovf$ccd_f)), color=set1[1])+
  geom_line(data=mcycle,aes(x=times,y=mean(odraws_gpcovf$ccd_f-2*odraws_gpcovf$ccd_sigma)), color=set1[1],
            linetype="dashed")+
  geom_line(data=mcycle,aes(x=times,y=mean(odraws_gpcovf$ccd_f+2*odraws_gpcovf$ccd_sigma)), color=set1[1],
            linetype="dashed")
```
### MCMC-based model

```{r}
mcycle %>%
  mutate(Ef=mean(draws_gpcovf$f),
         sigma=mean(draws_gpcovf$sigma)) %>%  
  ggplot(aes(x=times,y=accel))+
  geom_point()+
  labs(x="Time (ms)", y="Acceleration (g)")+
  geom_line(aes(y=Ef), color=set1[1])+
  geom_line(aes(y=Ef-2*sigma), color=set1[1],linetype="dashed")+
  geom_line(aes(y=Ef+2*sigma), color=set1[1],linetype="dashed")
```


## Numerical evaluation

```{r}
map_lengthscale_f_mean <- E(odraws_gpcovf$lengthscale_f)
map_sigma_f_mean <- E(odraws_gpcovf$sigma_f)
map_sigman_mean <- E(odraws_gpcovf$sigman)
map_gradient_evals <- 15
```


```{r}
ccd_lengthscale_f_mean <- weighted.mean(exp(ccd$points[,1]), ccd$weights)
ccd_sigma_f_mean <- weighted.mean(exp(ccd$points[,2]), ccd$weights)
ccd_sigman_mean <- weighted.mean(exp(ccd$points[,3]), ccd$weights)
ccd_gradient_evals <- sum(ccd$logden.count)
```

```{r}
mcmc_lengthscale_f_mean <- E(draws_gpcovf$lengthscale_f)
mcmc_sigma_f_mean <- E(draws_gpcovf$sigma_f)
mcmc_sigman_mean <- E(draws_gpcovf$sigman)
mcmc_gradient_evals <- sum(fit_gpcovf$sampler_diagnostics(format="df")$n_leapfrog__)
```

```{r}
results <- data.frame(
  "strategy" = c("MAP", "CCD", "MCMC"),
  "E(lengthscale_f)" = c(map_lengthscale_f_mean, ccd_lengthscale_f_mean, mcmc_lengthscale_f_mean),
  "E(sigma_f)" = c(map_sigma_f_mean, ccd_sigma_f_mean, mcmc_sigma_f_mean),
  "E(sigman)" = c(map_sigman_mean, ccd_sigman_mean, mcmc_sigman_mean),
  "gradient_evaluations" <- c(map_gradient_evals, ccd_gradient_evals, mcmc_gradient_evals)
)
results
```

## Convergence visualization

### Establishing the ground truth

```{r}
# model_gpcovf_large <- cmdstan_model(stan_file = "gpcovf.stan")
# fit_gpcovf_large <- model_gpcovf_large$sample(data=standata_gpcovf,
#                                   iter_warmup=500, iter_sampling=100000,
#                                   chains=4, parallel_chains=4, refresh=5000)
fit_gpcovf_large <- readRDS("fit_gpcovf_large")
```

```{r}
draws_gpcovf_large <- as_draws_rvars(fit_gpcovf_large$draws())
```

```{r}
true_lengthscale_f_mean <- E(draws_gpcovf_large$lengthscale_f)
true_sigma_f_mean <- E(draws_gpcovf_large$sigma_f)
true_sigman_mean <- E(draws_gpcovf_large$sigman)
```

### Calculate MAP and CCD errors

```{r}
map_lengthscale_f_error <- abs(map_lengthscale_f_mean - true_lengthscale_f_mean)
ccd_lengthscale_f_error <- abs(ccd_lengthscale_f_mean - true_lengthscale_f_mean)


```


### Convergence visualization

```{r}
draws_lengthscale_f <- as_draws_df(draws_gpcovf)$lengthscale_f
mcmc_lengthscale_f_error <- abs(cummean(as_draws_df(draws_gpcovf)$lengthscale_f) - true_lengthscale_f_mean)
gradient_evals <- cumsum(fit_gpcovf$sampler_diagnostics(format="df")$n_leapfrog__)
error_data <- data.frame(error=mcmc_lengthscale_f_error, gradient_evals=gradient_evals)

beats <- mcmc_lengthscale_f_error > ccd_lengthscale_f_error
evals_to_match_ccd <- gradient_evals[max(which(beats))]
beats <- mcmc_lengthscale_f_error > map_lengthscale_f_error
evals_to_match_map <- gradient_evals[max(which(beats))]
annotation_line_type <- 'dotted'
annotation_line_size <- 0.55

xl <- c(0, tail(gradient_evals, 1)-1000)
ggplot(error_data, aes(x=gradient_evals, y=error)) +
  geom_line(size=0.65, col="grey15") +
  annotate("text", x=tail(gradient_evals, 1) - 1000, y=0.0023, label="MCMC error", col='grey15', fontface='italic') +
  geom_hline(aes(yintercept=ccd_lengthscale_f_error), linetype=annotation_line_type, col='dodgerblue4', size=annotation_line_size) +
  annotate("text", x=tail(gradient_evals, 1) - 2000, y=ccd_lengthscale_f_error + 0.001, label="CCD error", col='dodgerblue4', fontface='italic') +
  geom_vline(aes(xintercept=evals_to_match_ccd), linetype=annotation_line_type, col='dodgerblue4', size=annotation_line_size) +
  annotate("text", x=evals_to_match_ccd + 900, y=0.022, label=sprintf("%d", evals_to_match_ccd), col='dodgerblue4', fontface='italic') +
  geom_hline(aes(yintercept=map_lengthscale_f_error), linetype=annotation_line_type, col='lightsteelblue4', size=annotation_line_size) +
  annotate("text", x=tail(gradient_evals, 1) - 4000, y=map_lengthscale_f_error + 0.001, label="MAP error", col='lightsteelblue4', fontface='italic') +
  geom_vline(aes(xintercept=evals_to_match_map), linetype=annotation_line_type, col='lightsteelblue4', size=annotation_line_size) +
  annotate("text", x=evals_to_match_map + 900, y=0.025, label=sprintf("%d", evals_to_match_map), col='lightsteelblue4', fontface='italic') +
  labs(y=TeX(r"( absolute error in $l_p$ )"), x="posterior gradient evaluations") +
  theme_bw()
```


# Reduced data set experiment

## Obtaining the data

```{r}
# reduced data set
mcycle_10p <- mcycle[seq(1,133,by=10),]
standata_gpcovf10p <- list(x=mcycle_10p$times,
                     x2=mcycle$times,
                     y=mcycle_10p$accel,
                     N=length(mcycle_10p$times),
                     N2=length(mcycle$times))
json_file <- toJSON(standata_gpcovf10p)
write(json_file, "gpcovf10p.json")
```

## MCMC sampling 

```{r}
model_gpcovf10p <- cmdstan_model(stan_file = "gpcovf.stan")
fit_gpcovf10p <- model_gpcovf10p$sample(data=standata_gpcovf10p,
                                  iter_warmup=500, iter_sampling=1000,
                                  chains=4, parallel_chains=4, refresh=1000)
```


```{r}
draws_gpcovf10p <- as_draws_rvars(fit_gpcovf10p$draws())
summarise_draws(subset(draws_gpcovf10p,
                       variable=c('sigma_','lengthscale_','sigma.'),
                       regex=TRUE))
```

## Fitting the model with bridgestan

```{r}
model10p <- StanModel$new("gpcovf_model10p.so", "gpcovf10p.json", 1234)
```

```{r}
print(paste0("This model's name is ", model10p$name(), "10p."))
print(paste0("This model has ", model10p$param_num(), " parameters."))
print(paste0("This model has parameters [", toString(model10p$param_unc_names()), "]."))
```

## Conducting the CCD point generation

```{r}
logdensity10p <- function(theta) {
  model10p$log_density(theta, jacobian=TRUE)
}
logdensity_gradient10p <- function(theta) {
  model10p$log_density_gradient(theta,  jacobian = TRUE)$gradient
}
logdensity_hessian10p <- function(theta) {
  model10p$log_density_hessian(theta,  jacobian = TRUE)$hessian
}

source('CCD_point_generation.r')
ccd10p <- CCD(logdensity10p,
           3,
           th0=c(-1,  0, -1),
           logden.grad=logdensity_gradient10p,
           logden.hessian=logdensity_hessian10p,
           optim.method='BFGS',
           auto.scale = T)
```

## Some visualizations for validation

```{r}
xl <- c(-0.1,2)
yl <- c(-0.3,4)
icons <- c(rep(22, 8), rep(21, 7))
colors <- c(rep('steelblue', 14), 'red')

as.data.frame(exp(ccd10p$points)) %>% map_df(rev) %>%
  ggplot(aes(x=V1, y=V2)) +
    coord_cartesian(xlim = xl, ylim = yl) +
    labs(x="lengthscale_f", y="sigma_f") +
    geom_point(data=as_draws_df(draws_gpcovf10p), aes(x=lengthscale_f, y=sigma_f), col=alpha('darkgrey', 0.1), pch=19) +
    geom_point(pch=icons, col=colors, size=2, stroke=1) +
    theme_bw()

xl <- c(-0.25,2)
as.data.frame(exp(ccd10p$points)) %>% map_df(rev) %>%
  ggplot(aes(x=V3, y=V2)) +
    coord_cartesian(xlim = xl, ylim = yl) +
    labs(x="sigman", y="sigma_f") +
    geom_point(data=as_draws_df(draws_gpcovf10p), aes(x=sigman, y=sigma_f), col=alpha('darkgrey', 0.15), pch=16) +
    geom_point(pch=icons, col=colors, size=2, stroke=1) +
    theme_bw()
```

## Optimized model

```{r}
opt_gpcovf10p <- model_gpcovf10p$optimize(data=standata_gpcovf10p,
                                    init=0.1, algorithm='bfgs')
odraws_gpcovf10p <- as_draws_rvars(opt_gpcovf10p$draws())
mode10p <- c(E(odraws_gpcovf10p$lengthscale_f), E(odraws_gpcovf10p$sigma_f), E(odraws_gpcovf10p$sigman))
subset(odraws_gpcovf10p, variable=c('sigma_','lengthscale_','sigma.'), regex=TRUE)
```


## Predictive plots

### Optimized model

```{r}
colors <- rep(alpha('black', 0.3), 133)
colors[seq(1,133,by=10)] <- alpha('blue3', 1)

mcycle %>%
  ggplot(aes(x=times,y=accel))+
  geom_point(col=colors)+
  labs(x="Time (ms)", y="Acceleration (g)", title="")+
  geom_line(data=mcycle,aes(x=times,y=mean(odraws_gpcovf10p$f)), color=set1[1])+
  geom_line(data=mcycle,aes(x=times,y=mean(odraws_gpcovf10p$f-2*odraws_gpcovf10p$sigma)), color=set1[1],
            linetype="dashed")+
  geom_line(data=mcycle,aes(x=times,y=mean(odraws_gpcovf10p$f+2*odraws_gpcovf10p$sigma)), color=set1[1],
            linetype="dashed")
```

### CCD model

```{r}
mcycle %>%
  ggplot(aes(x=times,y=accel))+
  geom_point(col=colors)+
  labs(x="Time (ms)", y="Acceleration (g)", title="")+
  geom_line(data=mcycle,aes(x=times,y=mean(odraws_gpcovf10p$ccd10p_f)), color=set1[1])+
  geom_line(data=mcycle,aes(x=times,y=mean(odraws_gpcovf10p$ccd10p_f-2*odraws_gpcovf10p$ccd10p_sigma)), color=set1[1],
            linetype="dashed")+
  geom_line(data=mcycle,aes(x=times,y=mean(odraws_gpcovf10p$ccd10p_f+2*odraws_gpcovf10p$ccd10p_sigma)), color=set1[1],
            linetype="dashed")
```

### MCMC-based model

```{r}
mcycle %>%
  mutate(Ef=mean(draws_gpcovf10p$f),
         sigma=mean(draws_gpcovf10p$sigma)) %>%  
  ggplot(aes(x=times,y=accel))+
  geom_point(col=colors)+
  labs(x="Time (ms)", y="Acceleration (g)")+
  geom_line(aes(y=Ef), color=set1[1])+
  geom_line(aes(y=Ef-2*sigma), color=set1[1],linetype="dashed")+
  geom_line(aes(y=Ef+2*sigma), color=set1[1],linetype="dashed")
```



## Numerical evaluation

```{r}
map10p_lengthscale_f_mean <- E(odraws_gpcovf10p$lengthscale_f)
map10p_sigma_f_mean <- E(odraws_gpcovf10p$sigma_f)
map10p_sigman_mean <- E(odraws_gpcovf10p$sigman)
map10p_gradient_evals <- 23
```


```{r}
ccd10p_lengthscale_f_mean <- weighted.mean(exp(ccd10p$points[,1]), ccd10p$weights)
ccd10p_sigma_f_mean <- weighted.mean(exp(ccd10p$points[,2]), ccd10p$weights)
ccd10p_sigman_mean <- weighted.mean(exp(ccd10p$points[,3]), ccd10p$weights)
ccd10p_gradient_evals <- sum(ccd10p$logden.count)
```

```{r}
mcmc10p_lengthscale_f_mean <- E(draws_gpcovf10p$lengthscale_f)
mcmc10p_sigma_f_mean <- E(draws_gpcovf10p$sigma_f)
mcmc10p_sigman_mean <- E(draws_gpcovf10p$sigman)
mcmc10p_gradient_evals <- sum(fit_gpcovf10p$sampler_diagnostics(format="df")$n_leapfrog__)
```

```{r}
results10p <- data.frame(
  "strategy" = c("MAP", "CCD", "MCMC"),
  "E(lengthscale_f)" = c(map10p_lengthscale_f_mean, ccd10p_lengthscale_f_mean, mcmc10p_lengthscale_f_mean),
  "E(sigma_f)" = c(map10p_sigma_f_mean, ccd10p_sigma_f_mean, mcmc10p_sigma_f_mean),
  "E(sigman)" = c(map10p_sigman_mean, ccd10p_sigman_mean, mcmc10p_sigman_mean),
  "gradient_evaluations" <- c(map10p_gradient_evals, ccd10p_gradient_evals, mcmc10p_gradient_evals)
)
results10p
```

# Extra stuff

## 

